Bandit Multiclass Linear Classification: Efficient Algorithms for the Separable Case

### 2019-01-23 ICML-submission/appendix-nearest-neighbor-algorithm.tex
Fixes

\begin{lemma}[Separation lemma]
\label{lemma:separation-lemma}
Let $(V,\ip{\cdot}{\cdot})$ be a vector space, $K$ be a positive integer and let
$\gamma$ be a positive real number. Suppose $(x_1,y_1), (x_2,y_2), \dots, (x_T,
y_T) \in V \times \{1,2,\dots,K\}$ are labeled examples that are weakly linearly
separable with margin $\gamma$. For $i$, $j$ in $\cbr{1,2,\dots,T}$, if
$\norm{x_i - x_j}_2 \le \gamma$ then $y_i = y_j$.
\end{lemma}

\begin{proof}
Suppose for the sake on contradiction that $y_i \neq y_j$. By
Definition~\ref{definition:linear-separability}, there exists a set of $k$
vectors $(w_1, \ldots, w_k)$, such that
conditions~\eqref{equation:weak-linear-separability-1}
and~\eqref{equation:weak-linear-separability-2} are satisfied.

@@

On the other hand,
$$
\ip{w_{y_i} - w_{y_j}}{x_i - x_j} \le \norm{w_{y_i} - w_{y_j}} \cdot \norm{x_i - x_j} < \sqrt{2} \gamma
$$
where the first inequality is from Cauchy-Schwartz inequality, the second
inequality is from that $\norm{w_{y_i} - w_{y_j}} \le \sqrt{2(\norm{w_{y_i}}^2 +
\norm{w_{y_j}}^2)} \leq \sqrt{2}$ and our assumption on $x_i$ and $x_j$.
Therefore, we reach a contradiction.
\end{proof}

We also need to define several notions. A subset $S \subseteq \R^d$ is called a
$\gamma$-packing if for any $x,x' \in S$ such that $x \neq x'$ we have $\norm{x -
x'} > \gamma$. The following lemma is standard. Also recall that $B(x,R) = \{ x'
\in \R^d ~:~ \norm{x' - x} \le R \}$ denotes the closed ball of radius $R$
centered a point $x$.

\begin{lemma}[Size of $\gamma$-packing]
\label{lemma:size-of-packing}
Let $\gamma$ and $R$ be positive real numbers.
If $S \subseteq B(\zero,R) \subseteq \R^d$ is a $\gamma$-packing then
$$
|S| \le \left( \frac{2R}{\gamma} + 1 \right)^d \; .
$$
\end{lemma}

\begin{proof}
If $S$ is a $\gamma$-packing then $\{ B(x,\gamma/2) ~:~ x \in S \}$
is a collection of disjoint balls of radius $\gamma$ that fit into $B(\zero,R + \gamma/2)$.
Thus,
$$
|S| \cdot \Vol(B(\zero, \gamma/2)) \le \Vol(B(\zero,R + \gamma/2))
$$
Hence,
\begin{align*}
|S|
& \le \frac{\Vol(B(\zero,R + \gamma/2))}{\Vol(B(\zero, \gamma/2))} \\
& = \left( \frac{R + \gamma/2}{\gamma/2} \right)^d \\
& = \left( \frac{2R}{\gamma} + 1 \right)^d \; .
\end{align*}
\end{proof}

@@

\begin{proof}
Let $M$ be the number of mistakes made by the algorithm. Let $b_t$ be the
indicator that line~\ref{line:nearest-neighbor-explore} is executed at timestep
$t$, i.e. we fall into the ``else'' case. Note that if $b_t = 0$, then by
Lemma~\ref{lemma:separation-lemma}, the prediction $\widehat{y}_t$ must equal
$y_t$, i.e. $z_t = 0$. Therefore, $M = \sum_{t=1}^T z_t = \sum_{t=1}^T b_t z_t$.
Let $U = \sum_{t=1}^T b_t (1-z_t)$. Clearly, $|S| = U$. Since $S \subseteq B(\zero, R)$
is a $\gamma$-packing, $U = |S| \le (\frac{2R}{\gamma} + 1)^d$.

>>>>

\begin{lemma}[Non-separation lemma]
\label{lemma:non-separation-lemma}
Let $(V,\ip{\cdot}{\cdot})$ be a vector space, $K$ be a positive integer and let
$\gamma$ be a positive real number. Suppose $(x_1,y_1), (x_2,y_2), \dots, (x_T,
y_T) \in V \times \{1,2,\dots,K\}$ are labeled examples that are weakly linearly
separable with margin $\gamma$. For $i$, $j$ in $\cbr{1,2,\dots,T}$, if
$\norm{x_i - x_j}_2 \le \gamma$ then $y_i = y_j$.
\end{lemma}

\begin{proof}
Suppose for the sake on contradiction that $y_i \neq y_j$. By
Definition~\ref{definition:linear-separability}, there exists
vectors $w_1, \ldots, w_K$ such that
conditions~\eqref{equation:weak-linear-separability-1}
and~\eqref{equation:weak-linear-separability-2} are satisfied.

@@

On the other hand,
$$
\ip{w_{y_i} - w_{y_j}}{x_i - x_j} \le \norm{w_{y_i} - w_{y_j}} \cdot \norm{x_i - x_j} \le \sqrt{2} \gamma
$$
where the first inequality is from Cauchy-Schwartz inequality, the second
inequality is from that $\norm{w_{y_i} - w_{y_j}} \le \sqrt{2(\norm{w_{y_i}}^2 +
\norm{w_{y_j}}^2)} \leq \sqrt{2}$ and our assumption on $x_i$ and $x_j$.
Therefore, we reach a contradiction.
\end{proof}

We also need to define several notions. A subset $S \subseteq \R^d$ is called a
$\gamma$-packing if for any $x,x' \in S$ such that $x \neq x'$ we have $\norm{x -
x'} > \gamma$. The following lemma is standard. Also recall that $\B(x,R) = \{ x'
\in \R^d ~:~ \norm{x' - x} \le R \}$ denotes the closed ball of radius $R$
centered a point $x$.

\begin{lemma}[Size of $\gamma$-packing]
\label{lemma:size-of-packing}
Let $\gamma$ and $R$ be positive real numbers.
If $S \subseteq \B(\zero,R) \subseteq \R^d$ is a $\gamma$-packing then
$$
|S| \le \left( \frac{2R}{\gamma} + 1 \right)^d \; .
$$
\end{lemma}

\begin{proof}
If $S$ is a $\gamma$-packing then $\{ \B(x,\gamma/2) ~:~ x \in S \}$
is a collection of disjoint balls of radius $\gamma$ that fit into $\B(\zero,R + \gamma/2)$.
Thus,
$$
|S| \cdot \Vol(\B(\zero, \gamma/2)) \le \Vol(\B(\zero,R + \gamma/2))
$$
Hence,
\begin{align*}
|S|
& \le \frac{\Vol(\B(\zero,R + \gamma/2))}{\Vol(\B(\zero, \gamma/2))} \\
& = \left( \frac{R + \gamma/2}{\gamma/2} \right)^d \\
& = \left( \frac{2R}{\gamma} + 1 \right)^d \; .
\end{align*}
\end{proof}

@@

\begin{proof}
Let $M$ be the number of mistakes made by the algorithm. Let $b_t$ be the
indicator that line~\ref{line:nearest-neighbor-explore} is executed at timestep
$t$, i.e. we fall into the ``else'' case. Note that if $b_t = 0$, then by
Lemma~\ref{lemma:non-separation-lemma}, the prediction $\widehat{y}_t$ must equal
$y_t$, i.e. $z_t = 0$. Therefore, $M = \sum_{t=1}^T z_t = \sum_{t=1}^T b_t z_t$.
Let $U = \sum_{t=1}^T b_t (1-z_t)$. Clearly, $|S| = U$. Since $S \subseteq \B(\zero, R)$
is a $\gamma$-packing, $U = |S| \le (\frac{2R}{\gamma} + 1)^d$.

### 2019-01-23 ICML-submission/mistake-lower-bound-for-ignorant-algorithms.tex
Fixes

An ignorant algorithm does not learn from mistakes in the following sense.
Suppose that an ignorant algorithm predicts incorrectly at time $t$ ($\widehat
y_t \neq y_t$). Then at time $t+1$, if it is given the same feature vector
$x_{t+1}=x_t$, it will predict the label with the same distribution
$p_{t+1}(\cdot|x_{t+1})=p_{t+1}(\cdot|x_t)=p_t(\cdot|x_t)$. Below we describe
the difficulty of learning from mistakes using a class of natural approaches
under the weak separability assumption. Consider an approach similar to OBAMA:
in each round $t$, we create a \textit{convex} loss function $\widehat \ell_t:
\R^{K\times d} \to \R$ that gives a loss value $\widehat
\ell_t(W)$ to all linear models $W \in \R^{K\times d}$ ($W=[w_1 \ldots
w_K]^\top$ for $w_1, \ldots, w_K$ defined in
Definition~\ref{definition:linear-separability}). The requirement for
$\widehat{\ell}_t$ to be convex is due to computational efficiency. To let
$\widehat{\ell}_t$ reflect the loss, if $\widehat y_t \neq y_t$, we want to
assign larger loss values $\widehat{\ell_t}(W)$ to the inconsistent models
$\mathcal{W}=\{W: \ip{w_{\widehat y_t}}{x_t} \ge \ip{w_i}{x_t}, \forall
i\in[K]\}$, while for other $W$'s, we want $\widehat{\ell_t}(W)$ to be smaller.
The following fact can be verified when $K\ge 3$, which suggests there is no
way to define the loss $\widehat \ell_t$ that satisfies the above properties
constraints: there exists $W_1, W_2, W_3$, where $W_2$ is in the line segment
between $W_1$ and $W_3$, such that $W_2\in \mathcal{W}$, while $W_1, W_3 \notin
\mathcal{W}$. On one hand, by convexity, we should have
$\widehat{\ell}_t(W_2)\le \max\{\widehat{\ell}_t(W_1),
\widehat{\ell}_t(W_3)\}$; on the other hand, to penalize $\mathcal{W}$, we
should have $\widehat{\ell}_t(W_2) > \widehat{\ell}_t(W_1)$ and
$\widehat{\ell}_t(W_2) > \widehat{\ell}_t(W_3)$. Clearly, these two requirements
contradict each other.

>>>>

An ignorant algorithm does not learn from mistakes in the following sense.
Suppose that an ignorant algorithm predicts incorrectly at time $t$ ($\widehat
y_t \neq y_t$). Then at time $t+1$, if it is given the same feature vector
$x_{t+1}=x_t$, it will predict the label with the same distribution
$p_{t+1}(\cdot|x_{t+1})=p_{t+1}(\cdot|x_t)=p_t(\cdot|x_t)$. Below we describe
the difficulty of learning from mistakes using a class of natural approaches
under the weak separability assumption. Consider an approach similar to OBAMA:
in each round $t$, we create a \textit{convex} loss function $\widehat \ell_t:
\R^{K\times d} \to \R$ that gives a loss value $\widehat
\ell_t(W)$ to all linear models $W \in \R^{K\times d}$ ($W=[w_1 \dots
w_K]^\top$ for $w_1, w_2, \dots, w_K$ defined in
Definition~\ref{definition:linear-separability}). The requirement for
$\widehat{\ell}_t$ to be convex is due to computational efficiency. To let
$\widehat{\ell}_t$ reflect the loss, if $\widehat y_t \neq y_t$, we want to
assign larger loss values $\widehat{\ell_t}(W)$ to the inconsistent models
$\mathcal{W}=\{W: \ip{w_{\widehat y_t}}{x_t} \ge \ip{w_i}{x_t}, \forall
i\in[K]\}$, while for other $W$'s, we want $\widehat{\ell_t}(W)$ to be smaller.
The following fact can be verified when $K\ge 3$, which suggests there is no
way to define the loss $\widehat \ell_t$ that satisfies the above properties
constraints: there exists $W_1, W_2, W_3$, where $W_2$ is in the line segment
between $W_1$ and $W_3$, such that $W_2\in \mathcal{W}$, while $W_1, W_3 \notin
\mathcal{W}$. On one hand, by convexity, we should have
$\widehat{\ell}_t(W_2)\le \max\{\widehat{\ell}_t(W_1),
\widehat{\ell}_t(W_3)\}$; on the other hand, to penalize $\mathcal{W}$, we
should have $\widehat{\ell}_t(W_2) > \widehat{\ell}_t(W_1)$ and
$\widehat{\ell}_t(W_2) > \widehat{\ell}_t(W_3)$. Clearly, these two requirements
contradict each other.
